---
title: "Normal Dynamic Linear Models"
author: "jagofc"
date: "18/04/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The general NDLM form

The class of Normal Dynamic Linear Models can be written as a pair of equations, the:

**observation equation:**  
$y_t = F_t' \theta_t + \nu_t$ with $\nu_t \sim N(0, v_t)$

**evolution equation (/system eq.):**  
$\theta_t = G_t \theta_{t-1} + \omega_t$ with $\omega_t \sim N(0, W_t)$.

Here:

+ $F_t$ is vector of dim $k$, assumed known
+ $G_t$ is $k$-by-$k$ matrix, assumed known
+ $v_t$ is the variance at the obs level, assumed known.
+ $W_t$ is the $k$-by-$k$ covariance matrix at the sys level, assumed known.
+ $\theta_t$ is a vector of dimension $k$ of parameters, to find.
+ $\theta_0 \sim N(m_0, c_0)$, assumed known.


The short-hand notation for this model is $\lbrace F_t, G_t, v_t, W_t \rbrace$.  
*Note: this ignores the prior on $\theta_0$*. 

We're interested in: 

+ filtering: $p(\theta_t | D_t)$
+ forecasting: $p(y_{t+h} | D_t)$
+ smoothing: $p(y_t | D_T)$ where $t < T$

where $D_t = \lbrace D_0, y_{1:t} \rbrace$ and $D_0$ is the information available at $t=0$.

We're also interested in point forecasts as well as densities. The forecast function is:

$$f_t(h) = \mathbb{E}(y_{t+h} | D_t) = F_{t+h}' G_{t+h} \cdots G_{t+1} \mathbb{E}(\theta_t | D_t)$$
and if $F_t = F$ and $G_t = G$ then we have:

$$f_t(h) = F'G^h \mathbb{E}(\theta_t| D_t)$$


# Polynomial trend model

+ First-order polynomial model: this has the form




+ Second-order polynomial model
+ $p$-th order polynomial model


# Regression models

$y_t = \beta_{t, 0} + \beta_{t, 1} X_t + \epsilon_t$ where $\epsilon_t \sim N(0, \nu_t)$

## General form

$y_t = \beta_{t, 0} + \beta_{t, 1} X_{t, 1} + \ldots + \beta_{t, p} X_{t, p} + \varepsilon_t$

$G = 2$
$F = 3$
$f_t(h) = 4$





```{r cars}
#summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```
